### 2014-11-07 contract-language.tex
edits
Edits
Added operational semantics to the techrep

\input{contract-syntax.tex}

@@

\caption{An Axiomatic Execution}
\label{sem:contracts}
\end{figure}

@@

  \vspace{-1.5em}
  \begin{smathpar}
  \scc = \forall a.~\sameobj{a}{\cureff} \Rightarrow \vis{a}{\cureff} ~\vee~ \vis{\cureff}{a} ~\vee~ a = \cureff
  \end{smathpar}

\end{itemize}

% Contract classification rules
% ------------------------------
\input{contract-classification}

@@

Our goal is to map application-level consistency constraints on operations to
appropriate store-level consistency guarantees capable of satisfying these
constraints.  The ability to express both these kinds of constraints as
contracts in our contract language lets us compare and determine if contract
($\cv_{op}$) of an operation ($\mathit{op}$) is weak enough to be satisfied
under a store consistency level identified by the contract $\cv_{st}$. Towards
this end, we define a binary \emph{weaker than} relation for our contract
language as following:
\begin{definition}
A contract $\cv_{op}$ is said to be weaker than $\cv_{st}$ (written $\cv_{op}
\le \cv_{st}$ ) if and only if $\Delta \vdash \cv_{st} \Rightarrow \cv_{op}$.
\begin{center}
\end{center}
\end{definition}
\vspace{-2em}
\noindent The $\Delta$ referred in the above defintion is a conjunction of
assumptions about the nature of primitive relations, e.g., $\Rvis$ and $\Rso$
are irreflexive, and the happens-before relation $\hbZ$ is acyclic. A
\emph{well-formed} axiomatic execution ($\E$) is expected to satisfy these
assumptions (i.e., $\E \models \Delta$).

@@

We now present a meta-theoretic result that certifies the soundness of
classification-based contract enforcement. To help us state the result, we
define an axiomatic model of the our system described informally in
\S~\ref{sec:sysmod}:

@@

We model the system as a tuple $\E,\Sigma$, where the axiomatic execution $\E$
captures the data store's current state, and session soup $\Sigma$ is the set
of concurrent client sessions interacting with the store. A session $\sigma$ is
a sequence of pairs composed of replicated data type operations $\mathit{op}$,
tagged with the consistency class $\tau$ of their contracts (as determined by
the contract classification scheme). We assume a rewrite relation of form:

@@

\noindent on the system state. The relation captures the progress of the
execution (from $\E$ to $\E'$)  due to the successful completion of a client
operation $\mathit{op}$ from one of the sessions in $\Sigma$, generating a new
effect $\eff$. If the resultant execution $\E'$ satisfies the store contract
$\cv_\tau$ (i.e., $\E \models \cv_\tau$), then we say that the store has
\emph{enforced} the contract $\cv_\tau$ in the execution $\E'$. With help of
the axiomatic model, we now state the soundness of contract enforcement as
follows:

@@

\noindent The theorem states that if a data store correctly enforces $\scc$,
$\ccc$, and $\ecc$ contracts in all well-formed executions, then the same
store, extended with the classification scheme shown in
Figure~\ref{sem:classify}, can enforce all well-formed \name contracts. The
proof of the theorem is given in the supplementary material\footnote{The
supplementary material also defines an operational semantics that concretizes
the rewrite relation ($\xrightarrow{}$), along with the proof that it correctly
enforces all \name contracts.}.

>>>>

\begin{figure}
\begin{smathpar}
\begin{array}{rclcl}
\multicolumn{5}{l}{
  {x,y,z} \in \mathtt{EffVar} \qquad
  {\cureff} \in \mathtt{CurEff} \qquad
  {\sf Op} \in \mathtt{OperName}
}\\
\cv 		& \in & \mathtt{Contract} 	& \coloneqq & \forall (x : \tau).\cv
        \ALT \pi \\
\tau		& \in	& \mathtt{EffType}	& \coloneqq &  {\sf Op}
        \ALT \tau \vee \tau \\
\pi			&	\in & \mathtt{Prop} & \coloneqq & \true \ALT R(x,y)
        \ALT \pi \vee \pi \\
			  & 		&	 &  \ALT & \pi \wedge \pi \ALT \pi \Rightarrow \pi \\
R				& \in & \mathtt{Relation}	& \coloneqq & \visZ \ALT \soZ
        \ALT \sameobjZ \ALT R^+ \\
				&			&	 &  \ALT & R \cup R \ALT R \cap R \\
\end{array}
\end{smathpar}
\caption{Contract language.}
\label{fig:contract-lang}
\end{figure}

@@

\caption{Axiomatic execution.}
\label{sem:contracts}
\end{figure}

@@

  \vspace{-1.5em}
  \begin{smathpar}
  \scc = \forall a.~\sameobj{a}{\cureff} \Rightarrow \vis{a}{\cureff} ~\vee~ \vis{\cureff}{a} ~\vee~ a = \cureff
  \end{smathpar}
\end{itemize}
\vspace{-1em}

@@

\newcommand{\DDe}[1]{#1}
\begin{figure}
\begin{smathpar}
\begin{array}{c}
\vspace{3mm}
\RuleTwo
{\DDe{\cv} \le \DDe{\scc}}
{{\sf WellFormed}(\cv)}  \qquad

\RuleTwo
{\DDe{\cv} \le \DDe{\ecc}}
{{\sf EventuallyConsistent}(\cv)} \\

\vspace{3mm}
\RuleTwo
{\DDe{\cv} \not\le \DDe{\ecc}
\quad \DDe{\cv} \le \DDe{\ccc}}
{{\sf CausallyConsistent}(\cv)} \qquad

\RuleTwo
{\DDe{\cv} \not\le \DDe{\ccc}
\quad \DDe{\cv} \le \DDe{\scc}}
{{\sf StronglyConsistent}(\cv)}
\end{array}
\end{smathpar}
\vspace{-5mm}
\caption{Contract classification.}
\label{sem:classify}
\end{figure}

Our goal is to map application-level consistency constraints on operations to
appropriate store-level consistency guarantees capable of satisfying these
constraints.  The ability to express both these kinds of constraints as
contracts in our contract language lets us compare and determine if contract
($\cv_{op}$) of an operation ($\mathit{op}$) is weak enough to be satisfied
under a store consistency level identified by the contract $\cv_{st}$. Towards
this end, we define a binary \emph{weaker than} relation for our contract
language as following:

\begin{definition}
A contract $\cv_{op}$ is said to be weaker than $\cv_{st}$ (written $\cv_{op}
\le \cv_{st}$ ) if and only if $\Delta \vdash \cv_{st} \Rightarrow \cv_{op}$.
\end{definition}

\noindent The $\Delta$ referred in the above defintion is a conjunction of
assumptions about the nature of primitive relations, e.g., $\Rvis$ and $\Rso$
are irreflexive, and the happens-before relation $\hbZ$ is acyclic. A
\emph{well-formed} axiomatic execution ($\E$) is expected to satisfy these
assumptions (i.e., $\E \models \Delta$).

@@

We now present a meta-theoretic result that certifies the soundness of
classification-based contract enforcement. To help us state the result, we
define an operational semantics of the our system described informally in
\S~\ref{sec:sysmod}:

@@

We model the system as a tuple $\E,\Sigma$, where the axiomatic execution $\E$
captures the data store's current state, and session soup $\Sigma$ is the set
of concurrent client sessions interacting with the store. A session $\sigma$ is
a sequence of pairs composed of replicated data type operations $\mathit{op}$,
tagged with the consistency class $\tau$ of their contracts (as determined by
the contract classification scheme). We assume a reduction relation of form:

@@

\noindent on the system state. The relation captures the progress of the
execution (from $\E$ to $\E'$)  due to the successful completion of a client
operation $\mathit{op}$ from one of the sessions in $\Sigma$, generating a new
effect $\eff$. If the resultant execution $\E'$ satisfies the store contract
$\cv_\tau$ (i.e., $\E \models \cv_\tau$), then we say that the store has
\emph{enforced} the contract $\cv_\tau$ in the execution $\E'$. With help of
the operational semantics, we now state the soundness of contract enforcement
as follows:

@@

\noindent The theorem states that if a data store correctly enforces $\scc$,
$\ccc$, and $\ecc$ contracts in all well-formed executions, then the same
store, extended with the classification scheme shown in
Figure~\ref{sem:classify}, can enforce all well-formed \name contracts. The
proof of the theorem is given below:

\begin{proof}
  Hypothesis:
  \begin{smathpar}
  \begin{array}{cr}
    \auxred{} {\E,\langle op,\tau \rangle;\sigma \pll \Sigma} {\eff}
    {\E', \sigma \pll \Sigma} & H\npp\\
    \E' \models \cv_\tau[\eff/\cureff] & H\npp\\
  \end{array}
  \end{smathpar}
  Since $\tau$ is the contract class of $\cv$, by inversion, we have
  $\cv \le \cv_\tau$. By the definition of $\le$ relation:
  \begin{smathpar}
  \begin{array}{cr}
    \Delta \vdash \forall \cureff.\,\cv_\tau \Rightarrow \cv & H\npp\\
  \end{array}
  \end{smathpar}
   Since $\eff$ denotes new effect, it is a fresh variable that does
   not occur free in $\Delta$. From $H2$, after instantiating bound
   $\cureff$ with $\eff$, we have:
  \begin{smathpar}
  \begin{array}{cr}
    \Delta \vdash \cv_\tau[\eff/\cureff] \Rightarrow \cv[\eff/\cureff]
      & H\npp\\
  \end{array}
  \end{smathpar}
  Due to the soundness of natural deduction for first-order logic,
  $H3$ implies that for all models $\mathcal{M}$ such that
  $\mathcal{M} \models \Delta$, if $\mathcal{M} \models
  \cv_\tau[\eff/\cureff]$ then $\mathcal{M} \models
  \cv[\eff/\cureff]$. Since $\E'$ is well-formed, we have:
  \begin{smathpar}
  \begin{array}{cr}
    \E' \models \Delta & H\npp\\
  \end{array}
  \end{smathpar}
  Proof follows from $H1$ and $H4$.
  \qed
\end{proof}

It is important to note that Theorem~\ref{thm:classification-sound}
does not ascribe any semantics to the reduction relation
($\xrightarrow{}$). As such, it makes no assumptions about how the
store actually implements ${\sf ec}$, ${\sf cc}$ and ${\sf sc}$
guarantees. The specific implementation strategy is determined by the
operational semantics of the store, which \emph{defines} the reduction
relation for that particular store. The following section describes
operational semantics of the store used by the \name implementation.

### 2014-11-07 store-opsem.tex
Added operational semantics to the techrep

\section{Appendix}
\label{sec:appendix}

\subsection{Soundness of Contract Classification}

We restate theorem ~\ref{thm:classification-sound} and provide its proof
below:

\begin{theorem}[Soundness of Contract Enforcement]
\label{lem:classification-sound}
Let $\cv$ be a well-formed contract of a replicated data type
operation $\mathit{op}$, and let $\tau$ denote the consistency class
of $\cv$ as determined by the contract classification scheme. For all
well-formed execution states $\E$, $\E'$ such that $\auxred{}
{\E,\langle op,\tau \rangle;\sigma \pll \Sigma} {\eff} {\E', \sigma
\pll \Sigma}$, if $\E' \models \cv_\tau[\eff/\cureff]$, then $\E'
\models \cv[\eff/\cureff]$
\end{theorem}
\begin{proof}
  Hypothesis:
  \begin{smathpar}
  \begin{array}{cr}
    \auxred{} {\E,\langle op,\tau \rangle;\sigma \pll \Sigma} {\eff}
    {\E', \sigma \pll \Sigma} & H\npp\\
    \E' \models \cv_\tau[\eff/\cureff] & H\npp\\
  \end{array}
  \end{smathpar}
  Since $\tau$ is the contract class of $\cv$, by inversion, we have
  $\cv \le \cv_\tau$. By the definition of $\le$ relation:
  \begin{smathpar}
  \begin{array}{cr}
    \Delta \vdash \forall \cureff.\,\cv_\tau \Rightarrow \cv & H\npp\\
  \end{array}
  \end{smathpar}
   Since $\eff$ denotes new effect, it is a fresh variable that does
   not occur free in $\Delta$. From $H2$, after instantiating bound
   $\cureff$ with $\eff$, we have:
  \begin{smathpar}
  \begin{array}{cr}
    \Delta \vdash \cv_\tau[\eff/\cureff] \Rightarrow \cv[\eff/\cureff]
      & H\npp\\
  \end{array}
  \end{smathpar}
  Due to the soundness of natural deduction for first-order logic,
  $H3$ implies that for all models $\mathcal{M}$ such that
  $\mathcal{M} \models \Delta$, if $\mathcal{M} \models
  \cv_\tau[\eff/\cureff]$ then $\mathcal{M} \models
  \cv[\eff/\cureff]$. Since $\E'$ is well-formed, we have:
  \begin{smathpar}
  \begin{array}{cr}
    \E' \models \Delta & H\npp\\
  \end{array}
  \end{smathpar}
  Proof follows from $H1$ and $H4$.
  \qed
\end{proof}

\subsection{Operational Semantics}
\renewcommand{\auxred}[4]{#1 \vdash #2 \;\xhookrightarrow{#3}\; #4 }

We now describe operational semantics of a data store that implements strong,
causal and eventual consistency guarantees. The semantics also serves as an
elegant representation of our implementation of \name.

Figure ~\ref{sem:oper} presents our semantics as concretization of the rewrite
relation ($\xrightarrow{}$) over execution state. Since we now have a concrete
store, we extend our system model with $\Theta$, a representation of the store
as a map from replicas to their local states. The local state of a replica $r$
(i.e., $\Theta(r)$) is a set of effects that are currently visible at $r$.  An
operation $op$ performed at replica $r$ can only witness the set of effects
($\Theta(r)$) visible at $r$. For the sake of clarity, we only consider a
single replicated object of well-defined type (for eg: a replicated object of
type \cf{BankAccount}) in our formalization.  Our semantics are parametric over
the specification of this replicated data type. Figure~\ref{sem:oper}
formalizes replicated data type (RDT) specification as tuple
$(\delta,\Ops,\Ctrts)$, where $\delta$ is the data type, $\Ops$ maps labels
($op$) of operations on $\delta$ to their definitions, while $\Ctrts$ maps them
to their consistency contracts ($\cv$). The definition of an operation is
expected to be a lambda expression, although we do not enforce this in our
formalization. For technical reasons, we tag each session with a session
identifier ($s$) and the sequence number ($i$) of the next operation in the
session.

The state of an operational execution ($\E$) is a tuple $\Exec$ where
$\EffSoup$ is a set of effects, and $\visZ,\soZ, \sameobjZ \subseteq \EffSoup
\times \EffSoup$ are \emph{visibility}, \emph{session order}, and \emph{same
object} relations over effects, respectively. We concretize an effect ($\eff$)
as a tuple $(s,i,op,v)$, which records the fact that $i^{th}$ action in session
with {\sf SessID} $s$, which is an operation $op$ on the replicated object, has
been successfully executed on some replica yielding a return value $v$. Note
that the combination of $s$ and $i$ uniquely identifies the effect. Session
order relation ($\soZ$) relates effects generated by the same session. An
effect $\eff = (s,i,op,v)$ is said to precede another effect $\eff' =
(s',i',op',v')$ in session order if and only if $s'=s$ and $i'\ge i$. Since we
only consider one replicated object in our formalization, the $\sameobjZ$
relation relates every pair of effects in the effect soup ($\EffSoup$). An
effect generated at a replica becomes visible at rest of the replicas
eventually.  If we denote the effect generated by the operation $op$ as
$\eff_{op}$, then $\Theta(r) \times \{\eff_{op}\} ~\subseteq~ \visZ$. Often, in
our formalization, we use $\visZ$ and $\soZ$ binary relations to obtain a set
of effects visible to a given effect $\eff$, or set of effects that precede a
given effect $\eff$ in the session order. As a syntactic convenience, whenever
$R$ is a binary relation, we write $R(\eff)$ to denote the set of all $\eff'$
such that $(\eff,\eff') \in R$.  Conversely, we write $R^{-1}(\eff)$ to denote
the set of all $\eff'$ such that $(\eff',\eff) \in R$.

@@

% Semantics
% ---------
\input{core-opsem}

@@

A {\sf StronglyConsistent} operation expects sequential consistency.
That is, universe of all effects ($\EffSoup$) in an execution ($\E$)
must be partitionable into a set of effects that \emph{happened
before} $\eff$ and another set that \emph{happened after} $\eff$,
where $\eff$ is the effect generated by an strongly consistent
operation. The rule $\rulelabel{UA}$ enforces this sequencing in two
steps; firstly, it insists that the the strongly consistent operation
($op$) witness effects of all operations executed so far by requiring
the global set of effects $\EffSoup$ to be a subset of local state
($\Theta(r)$) of the replica ($r$) executing $op$. Secondly, the rule
requires the effect ($\eff$) generated by $op$ to be added to the
local state of every other replica in the store, so that further
operations on these replicas can witness the effect of $op$. Since
both these steps require global coordination among replicas, the store
is \emph{unavailable} during the time it is executing $op$.
% However, it must be noted that executing strongly consistent
% operation does not entail the convergence of local states of all
% replicas; it simply means that there is one replica ($r$) whose
% local state is the upper bound of local states of all replicas, and
% there exists a set containing atleast one effect ($\eff$), which is
% their lower bound.

@@

Our main meta-theoretic result guarantees the soundness of our operational
semantics in enforcing $\ecc$, $\ccc$, and $\scc$ consistency guarantess by
ensuring that every reduction step preserves correctness of operational
execution.

\begin{theorem}[Soundness preservation]
\label{lem:core-preservation}
Let $\E$ be a well-formed operational execution state, and $\tau$ denote a
contract class.  If:
\begin{smathpar}
\E; \Theta; \langle s,i,\langle op,\tau \rangle;
\sigma \rangle \pll \Sigma \xhookrightarrow{\eff} \E'; \Theta';
\langle s,i,\sigma \rangle \Sigma'
\end{smathpar}
then (\romannumeral 1) $\E'$ is well-formed, and (\romannumeral 2)
$\E' \models \cv_{\tau}[\eff/\cureff]$
\end{theorem}

>>>>

\section{Operational Semantics}
\label{sec:store-opsem}

\renewcommand{\auxred}[4]{#1 \vdash #2 \;\xhookrightarrow{#3}\; #4 }

We now describe operational semantics of a data store that implements
strong, causal and eventual consistency guarantees. The semantics also
serves as a high-level description of our implementation of the store
underlying \name.

Figure ~\ref{sem:oper} presents operational semantics as rules
defining the reduction relation ($\xrightarrow{}$) over the execution
state.  Since we now have a concrete store, we extend our system model
with $\Theta$, a representation of the store as a map from replicas to
their local states. The local state of a replica $r$ (i.e.,
$\Theta(r)$) is a set of effects that are currently visible at $r$.
An operation $op$ performed at replica $r$ can only witness the set of
effects ($\Theta(r)$) visible at $r$. For the sake of clarity, we only
consider a single replicated object of well-defined type (for eg: a
replicated object of type \cf{BankAccount}) in our formalization.  Our
semantics are parametric over the specification of this replicated
data type. Figure~\ref{sem:oper} formalizes replicated data type (RDT)
specification as tuple $(\delta,\Ops,\Ctrts)$, where $\delta$ is the
data type, $\Ops$ maps labels ($op$) of operations on $\delta$ to
their definitions, while $\Ctrts$ maps them to their consistency
contracts ($\cv$). The definition of an operation is expected to be a
lambda expression, although we do not enforce this in our
formalization. For technical reasons, we tag each session with a
session identifier ($s$) and the sequence number ($i$) of the next
operation in the session.

The state of an operational execution ($\E$) is a tuple $\Exec$ where
$\EffSoup$ is a set of effects, and $\visZ$, $\soZ$, $\sameobjZ$
$\subseteq$ $\EffSoup \times \EffSoup$ are \emph{visibility},
\emph{session order}, and \emph{same object} relations over effects,
respectively. We concretize an effect ($\eff$) as a tuple
$(s,i,op,v)$, which records the fact that $i^{th}$ action in session
with {\sf SessID} $s$, which is an operation $op$ on the replicated
object, has been successfully executed on some replica yielding a
return value $v$. Note that the combination of $s$ and $i$ uniquely
identifies the effect. Session order relation ($\soZ$) relates effects
generated by the same session.  An effect $\eff = (s,i,op,v)$ is said
to precede another effect $\eff' = (s',i',op',v')$ in session order if
and only if $s'=s$ and $i'\ge i$. Since we only consider one
replicated object in our formalization, the $\sameobjZ$ relation
relates every pair of effects in the effect soup ($\EffSoup$). An
effect generated at a replica becomes visible at rest of the replicas
eventually.  If we denote the effect generated by the operation $op$
as $\eff_{op}$, then $\Theta(r) \times \{\eff_{op}\} ~\subseteq~
\visZ$. Often, in our formalization, we use $\visZ$ and $\soZ$ binary
relations to obtain a set of effects visible to a given effect $\eff$,
or set of effects that precede a given effect $\eff$ in the session
order. As a syntactic convenience, whenever $R$ is a binary relation,
we write $R(\eff)$ to denote the set of all $\eff'$ such that
$(\eff,\eff') \in R$.  Conversely, we write $R^{-1}(\eff)$ to denote
the set of all $\eff'$ such that $(\eff',\eff) \in R$.

@@

% Semantics
% ---------
\input{store-opsem-rules}

@@

A {\sf StronglyConsistent} operation expects sequential consistency.
That is, universe of all effects ($\EffSoup$) in an execution ($\E$)
must be partitionable into a set of effects that \emph{happened
before} $\eff$ and another set that \emph{happened after} $\eff$,
where $\eff$ is the effect generated by an strongly consistent
operation. The rule $\rulelabel{SC}$ enforces this sequencing in two
steps; firstly, it insists that the the strongly consistent operation
($op$) witness effects of all operations executed so far by requiring
the global set of effects $\EffSoup$ to be a subset of local state
($\Theta(r)$) of the replica ($r$) executing $op$. Secondly, the rule
requires the effect ($\eff$) generated by $op$ to be added to the
local state of every other replica in the store, so that further
operations on these replicas can witness the effect of $op$. Since
both these steps require global coordination among replicas, the store
is \emph{unavailable} during the time it is executing $op$.
% However, it must be noted that executing strongly consistent
% operation does not entail the convergence of local states of all
% replicas; it simply means that there is one replica ($r$) whose
% local state is the upper bound of local states of all replicas, and
% there exists a set containing atleast one effect ($\eff$), which is
% their lower bound.

@@

We now prove a meta-theoretic property that establishes the soundness
of our operational semantics in enforcing $\ecc$, $\ccc$, and $\scc$
consistency guarantees at every reduction step. As a corollary of this
result, and Theorem~\ref{thm:classification-sound}, we have the
assurance that \name correctly enforces all well-formed consistency
contracts.

\begin{theorem}[Soundness Modulo Classification]
\label{thm:weak-soundness}
For every well-formed execution state $\E$, and contract class $\tau 
\in \{{\sf ec},{\sf cc},{\sf sc}\}$, If:
\begin{smathpar}
(\E, \Theta, \langle s,i,\langle op,\tau \rangle;
\sigma \rangle \pll \Sigma) \xrightarrow{\eff} (\E', \Theta',
\langle s,i,\sigma \rangle \Sigma')
\end{smathpar}
then (\romannumeral 1) $\E'$ is well-formed, and (\romannumeral 2)
$\E' \models \cv_{\tau}[\eff/\cureff]$
\end{theorem}
\begin{proof}
  By case analysis on $(\E, \Theta, \langle s,i,\langle op,\tau \rangle;
  \sigma \rangle \pll \Sigma) \xrightarrow{\eff} (\E', \Theta',
  \langle s,i,\sigma \rangle \Sigma')$. 
  \GK{Lets first finalize all notations, rules, and also agree on
  all theorem statements. This way we don't have to change the proof
  again.} 
\end{proof}

\begin{corollary}[Soundness]
\label{thm:soundness}
For every well-formed execution state $\E$, and contract class $\tau
\in \{{\sf ec},{\sf cc},{\sf sc}\}$, and for every consistency
contract $\cv$ in the contract class $\tau$, If:
\begin{smathpar}
(\E, \Theta, \langle s,i,\langle op,\tau \rangle;
\sigma \rangle \pll \Sigma) \xrightarrow{\eff} (\E', \Theta',
\langle s,i,\sigma \rangle \Sigma')
\end{smathpar}
then (\romannumeral 1) $\E'$ is well-formed, and (\romannumeral 2)
$\E' \models \cv[\eff/\cureff]$
\end{corollary}
\begin{proof}
Follows from Theorems~\ref{thm:classification-sound} and
\ref{thm:weak-soundness}
\qed
\end{proof}
